---
title: "venny"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{venny}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

# 1. Overview

The `venny` package provides an integrated workflow for exploring relationships among sets with
Venn diagrams, summary tables, and polygon-based set operations.

Its functionality can be summarized as:

* **Venn diagram visualization** for 2 to 4 sets;
* **subset summaries**, including the number and percentage of elements in each region;
* **direct access to subset elements** for downstream analysis;
* **ellipse paths** underlying the Venn diagram, which can be reused as polygons;
* **set operations on polygon paths**, including intersection, union, and set difference;
* **highlighting selected regions** of an existing Venn diagram;
* **fine-grained control** over labels, positions, fonts, fills, lines, and transparency.

The package is particularly useful when a Venn diagram is not only the final visualization, but
also an intermediate object for subsequent set-based analysis.

<br>

# 2. Installation

Install the released version from CRAN:

```{r installation-cran, eval=FALSE}
install.packages("venny")
```

Or install the development version from GitHub:

```{r installation-github, eval=FALSE}
if (!require("pak")) install.packages("pak")
pak::pak("P10911004-NPUST/venny")
```

Then load the package:

```{r load-package}
library(venny)
```

<br>

# 3. Quick start

The simplest input is a named list containing two to four vectors.

```{r quick-start}
venny(LGL23$DEGs)
```

A practical usage scenario:

```{r}

```


The names of the list are used as the set labels. If the input list is unnamed, `venny`
automatically uses labels such as `Set_A`, `Set_B`, and `Set_C`.

```{r unnamed-input}
data <- list(1:100, 51:150)
venny(unname(LGL23$DEGs))
```

The basic plot is a `ggplot` object, so ordinary `ggplot2` layers can be added to it.

<br>

# 4. Understanding the subsets

For three sets, the possible subsets are represented by labels such as `A`, `AB`, `ABC`, and so on.

```{r}
venny(unname(LGL23$DEGs[-4]))
```

For example:

* `A` = elements belonging only to set A;
* `AB` = elements belonging to A and B but not C;
* `ABC` = elements shared by all three sets.

The helper `how_many_subsets()` can be used to determine how many subsets are possible.

```{r subset-number}
how_many_subsets(c("A", "B", "C"))
how_many_subsets(c("A", "B", "C"), detail = TRUE)
```

The corresponding binary representation can be generated with `bits_encoding()`:

```{r bits}
bits_encoding(c("A", "B", "C"))
```

This is useful when constructing or inspecting subset definitions programmatically.

<br>

# 5. Summary tables

When the numerical summary is more important than the diagram, use `venn_summary()`.

```{r summary}
out <- venn_summary(LGL23$DEGs)
out$table
```

The summary contains the information needed to describe each subset.

This makes `venn_summary()` useful as a bridge between visualization and downstream analysis.
For example, a specific subset can be extracted and passed to enrichment analysis, annotation, 
or another statistical workflow.

<br>

# 6. Obtaining detailed output

By default, `venny()` returns a Venn diagram. Set `detail = TRUE` to obtain the diagram together 
with the objects needed for downstream analysis.

```{r detail}
out <- venny(LGL23$DEGs, detail = TRUE)
names(out)
```

The returned object contains:

* `venn` — the Venn diagram as a `ggplot` object;
* `ellipse_path` — polygon coordinates for the individual ellipses;
* `table` — the subset summary table;
* `subset_elements` — the elements belonging to each subset.

For example:

```{r detail-components}
out$venn
```

The summary table can be inspected directly:

```{r detail-table}
out$table
```

And the elements belonging to a particular subset can be retrieved from:

```{r detail-elements}
out$subset_elements$ABC
```

<br>

# 7. Customizing the Venn diagram

`venny()` separates plot customization into several parameter functions. This allows the same 
parameter object to be reused across multiple diagrams.

## 7.1 Ellipse lines and fills

Use `ellipse_line()` and `ellipse_fill()` to control the appearance of the ellipses, *i.e.* set 
boundaries and interiors.

```{r ellipse-style}
venny(
    LGL23$DEGs,
    ellipse.line = ellipse_line(color = c("red", "transparent", "red", "navy"),
                                linetype = "solid", 
                                linewidth = 2, 
                                alpha = 0.8),
    ellipse.fill = ellipse_fill(color = c("gray", "red", "navy", "orange"),
                                alpha = 0.25)
)
```

The default palette is designed to provide distinct colors for up to four sets.

## 7.2 Set labels

Set labels are the names for each datasets. For example, the set labels of `LGL23$DEGs` are:

```{r}
names(LGL23$DEGs)
```

However, the set labels for internal recognition is fixed as `set_label_default()`. The positioning 
are:

| Number of sets  | Internal set labels | Positioning                        |
| :---:           | :--------:          | :----------:                       |
| 2               | A, B                | left -> right                      |
| 3               | A, B, C             | upper -> lower left -> lower right |
| 4               | A, B, C, D          | start from lower left, clockwise   |

So, when you want to assign the selected sets to the parameters-related function, do not use the 
user-defined set labels. For example, if you want to hide the set labels of "KO_low vs KO_mock" 
(lower left) and "OE_mock vs KO_mock" (upper right), you need to select "A" and "C".

```{r set-label-hide}
venny(
    LGL23$DEGs,
    set.label.position = set_label_position(
        hjust = c(0, -1, 0.5, 0),
        hide = c("A", "C")
    )
)
```

The appearance of set labels is controlled with `set_label_font()`:

```{r set-label-font}
venny(
    LGL23$DEGs,
    set.label.font = set_label_font(
        face = "italic",
        size = 7
    )
)
```

## 7.3 Subset labels

Subset labels can be customized via `subset_*_*()` functions independently of set labels. The subset
labels can be renamed by assigning a list of (old_name = new_name) to the `subset.label` argument.

```{r subset-label-rename}
subset_font_color <- vapply(
    subset_label_default(4), 
    function(x) if (x %in% c("A", "B", "C", "D")) "maroon" else "blue",
    FUN.VALUE = character(1)
)

venny(
    LGL23$DEGs,
    subset.label = list(
        A = "Apple",
        B = "Banana",
        C = "Coconut",
        D = "Durian"
    ),
    subset.label.font = subset_label_font(
        color = subset_font_color
    )
)
```

## 7.4 Counts and percentages

Counts and percentages are independent layers.

```{r count-percentage}
nm <- subset_label_default(4)
font_color <- fixed_length(c("red", "blue", "maroon"), length(nm))
font_angle <- fixed_length(c(30, 60, 90, 180), length(nm))
show_subset_perc <- c("A", "B", "C", "D", "AB", "ACD", "BC", "CD", "ABCD")

venny(
    LGL23$DEGs,
    subset.count = TRUE,
    subset.count.font = subset_count_font(family = "serif",
                                          color = font_color,
                                          angle = font_angle),
    subset.percentage = TRUE,
    subset.percentage.rounding = 1,
    subset.percentage.position = subset_percentage_position(show = show_subset_perc),
)
```


For example:

```{r count-font}
venny(
    LGL23$DEGs,
    subset.count.font = subset_count_font(
        face = "bold",
        size = 4
    ),
    subset.percentage.font = subset_percentage_font(
        size = 3.5
    )
)
```

<br>

# 8. Working with ellipse paths

One of the distinctive features of `venny` is that the Venn diagram is not treated only as a 
picture. The individual ellipses are returned as polygon coordinates.

```{r ellipse-path}
out <- venny(LGL23$DEGs, detail = TRUE)
names(out$ellipse_path)
```

Each element of `out$ellipse_path` contains the coordinates needed to draw the corresponding 
polygon. You can inspect one ellipse directly:

```{r one-ellipse}
head(out$ellipse_path$`KO_low vs KO_mock`)
```

The lower-level `generate_ellipse_path()` function can also be used independently.

```{r generate-ellipse}
circle <- generate_ellipse_path(
    x0 = 0,
    y0 = 0,
    a = 1,
    b = 1
)

head(circle)
```

Because the result is a data frame of `(x, y)` coordinates, it can be plotted with `ggplot2`:

```{r generate-ellipse-plot}
ggplot2::ggplot(
    circle,
    ggplot2::aes(x, y)
) +
    ggplot2::geom_polygon() +
    ggplot2::coord_fixed() +
    ggplot2::theme_void()
```

<br>

# 9. Set operations on Venn regions

The ellipse paths returned by `venny()` can be combined with set operations.

The package provides methods for:

* `intersect()`: {A} &cap; {B} &cap; ... -- the common elements shared by the sets;
* `union()`: {A} &cup; {B} &cup; ... -- all elements in the sets;
* `setdiff()`: {A} &minus; {B} &minus; ... -- subtraction from the previous sets.

These operations can subsequently be displayed with `highlight()` or `ggplot2::geom_polygon`. The 
`highlight()` is a convenient wrapper around `ggplot2::geom_polygon()` for displaying the result 
of a polygon operation.

## 9.1 Intersection

```{r polygon-intersection}
out <- venny(unname(LGL23$DEGs), detail = TRUE)
ep <- out$ellipse_path
setops <- venny::intersect(ep$Set_A, ep$Set_B, ep$Set_D)
highlight(out$venn, setops)
```

This highlights the region shared by A, B, and D. Because `highlight()` returns a `ggplot` object, 
additional layers can be added. For example

```{r highlight-annotation}
highlight(
    out$venn,
    setops,
    color = "red",
    fill = "red",
    alpha = 0.25
) +
    ggplot2::annotate(
        "text",
        x = 0,
        y = 2.5,
        label = "A ∩ B ∩ D",
        fontface = "bold"
    )
```

## 9.2 Union

```{r polygon-union}
setops <- venny::union(ep$Set_A, ep$Set_C, ep$Set_D)
highlight(out$venn, setops)
```

## 9.3 Set difference

```{r polygon-difference}
setops <- venny::setdiff(ep$Set_B, ep$Set_D)
highlight(out$venn, setops)
```

## 9.4 Chaining operations

Set operations can be chained to describe more complex regions.

```{r polygon-chain}
setops <- venny::union(ep$Set_A, ep$Set_B, ep$Set_C) |>
    venny::intersect(ep$Set_D) |>
    venny::setdiff(ep$Set_B)
highlight(out$venn, setops)
```

This approach is useful when the region of interest cannot be described simply by selecting a 
single named Venn subset.

<br>

# 10. RNA-seq example

This is not a fully realistic case study. The purpose is to demonstrate how to use this package, 
rather than to present a rigorous academic analysis.

## Prerequisites

```{r, message=FALSE}
if (!require(venny)) install.packages("venny")
if (!require(dplyr)) install.packages("dplyr")
if (!require(forcats)) install.packages("forcats")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(ggtext)) install.packages("ggtext")
if (!require(BiocManager)) install.packages("BiocManager")
if (!require(clusterProfiler)) BiocManager::install("clusterProfiler")
if (!require(org.At.tair.db)) BiocManager::install("org.At.tair.db")
library(venny)
library(dplyr)
library(forcats)
library(ggplot2)
library(ggtext)
library(clusterProfiler)
library(org.At.tair.db)

intersect <- venny::intersect
setdiff <- venny::setdiff
union <- venny::union
```

## Background information

```{r}
lst <- LGL23$DEGs
print(names(lst))
```

The `LGL23` object is an RNA-seq dataset stored as a list containing the following components:

- sample_info: A data frame containing the genotype and treatment information for each sample ID.
- GFD: A data frame containing gene functional descriptions and annotations.
- count_matrix: A matrix of read counts (gene expression) across all sample IDs.
- DEGs: A list of differentially expressed genes (DEGs). This is an example for illustration.
  - KO_low vs KO_mock: 
    Genes that are differentially expressed in the knockout (KO) plant after low-dose chemical 
    treatment compared with the untreated KO control. The KO plant is a genetically modified line 
    in which a specific gene (e.g., gene_X) has been disrupted or deleted.
  - KO_high vs KO_mock: 
    Similar to KO_low vs KO_mock, but the KO plant is treated with a high dose of the chemical.
  - WT_mock vs KO_mock: 
    Genes that are differentially expressed between the wild-type (WT) plant and the untreated KO 
    plant under mock-treatment conditions.
  - OE_mock vs KO_mock: 
    Genes that are differentially expressed between the overexpression (OE) plant and the untreated 
    KO plant under mock-treatment conditions. The OE plant is a genetically modified line in which 
    gene_X is highly expressed or constitutively activated. Ideally, the transcriptional profile of 
    OE_mock should resemble that of KO_high.

## Loss-of-function (Set C)

To investigate the function of the target gene, we compared gene expression profiles between the 
WT and KO plants under untreated conditions. Differentially expressed genes (DEGs) identified from 
this comparison may provide insights into the biological roles and regulatory functions of the 
target gene.

```{r, fig.width=8, fig.height=4}
subset_names <- names(subset_label_default(length(lst)))
select_subsets <- c("C", "BC", "CD", "ABC", "BCD", "AC", "ABCD", "ACD")
font_color <- sapply(subset_names, \(x) if (x %in% select_subsets) "red" else "navy")

out <- venny(
    data = lst,
    detail = TRUE,
    set.label.position = set_label_position(hjust = c(0.2, -0.5, 0.5, -0.2)),
    subset.label.font = subset_label_font(color = font_color),
    subset.count.font = subset_count_font(color = font_color),
    subset.percentage.font = subset_percentage_font(color = font_color)
)

venn <- out$venn
setops <- out$ellipse_path$`WT_mock vs KO_mock`

highlight(venn, setops, linetype = "solid", color = "red") +
    coord_cartesian(xlim = c(-3, 5)) +
    annotate("richtext",
        x = 3.2, y = -1.6,
        hjust = 0,
        size = 5,
        label = paste(
            "<b>KO:</b> Knock-out",
            "<b>WT:</b> Wild-type",
            "<b>OE:</b> Overexpression",
            "<b>mock:</b> 0 nM treatment",
            "<b>low:</b> 1 nM treatment",
            "<b>high:</b> 5 nM treatment",
            sep = "<br>"
        ),
    )
```

Target genes can be extracted from the DEGs and further analyzed using Gene Ontology (GO) analysis. 
The results suggest that the gene is involved in regulating nitrogen metabolism. Consistent with the 
finding, we observed clear phenotypic differences between WT and KO plants under nitrogen-deficient 
conditions. Additionally, responses under sulfate starvation and drought recovery reveal potential 
new functional roles of the gene, providing directions for future investigation.

```{r}
GO <- clusterProfiler::enrichGO(
    gene = LGL23$DEGs$`WT_mock vs KO_mock`,
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "BP"
)

GO@result |>
    slice_max(RichFactor, n = 20) |>
    ggplot(aes(RichFactor, fct_reorder(Description, RichFactor))) +
    theme_bw() +
    geom_point(aes(size = Count, color = FoldEnrichment)) +
    theme(axis.title.y = element_blank())
```

## High-dosage recovery (Subset BCD)

We compared transcriptomic responses induced by high-dose exogenous chemical treatment in the 
knockout line (KO_high), endogenous overexpression of the target gene in the OE line (OE_mock), 
and the wild-type control (WT_mock), each relative to the untreated knockout condition (KO_mock). 
This integrative comparison was used to evaluate whether the exogenous chemical can phenocopy 
endogenous activation of the pathway and whether its transcriptional effects converge toward a 
wild-type-like state.

```{r}
select_subsets <- "BCD"
font_color <- sapply(subset_names, \(x) if (x %in% select_subsets) "red" else "navy")

BCD <- venny(
    data = lst,
    detail = TRUE,
    set.label.position = set_label_position(hjust = c(0.2, -0.5, 0.5, -0.2)),
    subset.label.font = subset_label_font(color = font_color),
    subset.count.font = subset_count_font(color = font_color),
    subset.percentage.font = subset_percentage_font(color = font_color)
)

venn <- BCD$venn
ep <- BCD$ellipse_path
setops <- ep$`KO_high vs KO_mock` |>
    intersect(ep$`WT_mock vs KO_mock`, ep$`OE_mock vs KO_mock`) |>
    setdiff(ep$`KO_low vs KO_mock`)

highlight(venn, setops, linetype = "solid", color = "red")
```

```{r}
GO <- clusterProfiler::enrichGO(
    gene = BCD$subset_elements$BCD,
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "BP"
)

GO@result |>
    slice_max(RichFactor, n = 20) |>
    ggplot(aes(RichFactor, fct_reorder(Description, RichFactor))) +
    theme_bw() +
    geom_point(aes(size = Count, color = FoldEnrichment)) +
    theme(axis.title.y = element_blank())
```

Across all three contrasts, we observed a consistent enrichment of genes involved in nitrogen 
metabolism, suggesting that KO_high, OE_mock, and WT_mock share a common regulatory signature in 
this pathway. This result supports the hypothesis that the synthesized chemical functionally mimics 
the endogenous gene activity and partially restores wild-type-like nitrogen metabolic regulation 
in the KO background.

<br>

# 11. Practical recommendations

## Choose `venny()` when:

* you need a publication-oriented Venn diagram;
* you want control over labels, counts, percentages, fills, and lines;
* you need direct access to the underlying subset elements;
* you want to perform additional set operations on diagram regions.

## Choose `venn_summary()` when:

* the numerical composition of the subsets is more important than visualization;
* you need a table for downstream processing;
* you want to extract the elements belonging to each region.

## Use ellipse-path operations when:

* the desired region is more complicated than a single Venn subset;
* you need to combine intersections, unions, and differences;
* you want to highlight a custom polygonal region in a Venn diagram.

<br>

# Summary

The `venny` package provides three closely connected levels of analysis:

1. **Visualization** — `venny()` creates customizable Venn diagrams.
2. **Tabulation** — `venn_summary()` describes the composition of every subset and can expose the corresponding elements.
3. **Geometry and set operations** — ellipse paths can be manipulated with `intersect()`, `union()`, and `setdiff()`, and the resulting regions can be visualized with `highlight()`.

This makes `venny` useful not only for displaying overlaps, but also for constructing reproducible workflows in which Venn regions become explicit objects for downstream analysis.

---


<div style="text-align: center; margin-top: 2em; margin-bottom: 2em;">
<h2>The End</h2>
</div>
